Inner loop optimizations in mapping single-threaded programs to hardware

### 2013-03-27 Documents/Publications/dsd2013/paper.tex
Updated, ready to submit.

\author{Madhav P. Desai\\
  Indian Institute of Technology -- Bombay, Powai, Mumbai -- 400076, INDIA\\
  Email: madhav@ee.iitb.ac.in }
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

@@

In our work, we have developed a set of compiler tools which take a C-description of
an algorithm (with the only important restrictions being the requirement that
cycles are not present in the call graph of the source program, and the lack of 
support for {\bf function} pointers, normal pointers are supported), and produces
a VHDL description which is provably equivalent to the original program \cite{SameerPhD}.
Results on the application of this compiler to single-threaded programs show
that the circuits (when synthesized to VLSI layout) are substantially more
energy efficient than a processor, but that the performance of the processor
is usually better than that of the synthesized circuit \cite{dsd2010}.
If multi-threaded source programs are used as a starting point, the circuits produced 
are quite competitive with hand-designed RTL \cite{usenix2012}.
Quantitatively, the results in \cite{dsd2010} and \cite{usenix2012} indicate
that the gap between the synthesized hardware and modern
pipelined processors (for single-threaded
programs) and between the synthesized hardware and hand-crafted hardware (for
multi-threaded programs) is of the order of 2-3X.  That is, the performance
of the synthesized hardware needs to improve by 2-3X in order to close
the respective performance gaps.

@@

In the context of compilation to a pipelined processor, 
several loop optimizations
have been considered in literature, such as loop-unrolling, loop-peeling,
software-loop-pipelining etc. \cite{wolfe}, \cite{muchnick}, with the intent of
these optimizations being the extraction of as much
parallelism as possible from the single-threaded source program.

%% Algorithm-to-hardware:  can similar techniques work?
Similar loop-optimization techniques have been explored and
reported in the literature related to reconfigurable hardware (for
example \cite{Weinhardt}, \cite{Cardoso}, \cite{Kastner}).  For instance,
in the work reported by \cite{Weinhardt},
loop optimizations are done in a manner analogous to the static
techniques used in software compilers, in which index 
expressions which depend on the induction variable
are analysed to identify dependencies and schedule operations
across iterations of the loop body.  An explicitly timed controller is synthesized
for the pipeline.  Another approach which works in a similar
manner is described in \cite{Cardoso}.  These approaches rely on a static analysis of
the loop, and the cases to which the approach can be applied
are restricted (but are still sufficiently general for most linear
algebra and digital signal processing kernels).

@@

We give a brief description of our compiler flow.  The
details are not relevant to this paper, and the interested
reader can find them in \cite{dsd2010}.

@@

The compiler flow has been characterized over a wide variety
of applications \cite{dsd2010}.   The Click-to-NetFPGA tool-chain \cite{usenix2012} 
uses this compiler flow to map modular network appliances written in C++ to
an FPGA platform, and achieved about half the
performance possible using equivalent hand designed hardware.  
Thus,  the flow can be applied to ``real'' software
to produce competitive results.  The optimizations indicated in this
paper, should in principle, make the flow even more competitive.

@@

The virtual circuit generated by our compiler consists of three
cooperating components: the control-path, the data-path and
the storage system \cite{ahir}

To illustrate the model, we consider a simple example.
\begin{verbatim}
float a[1024], b[1024];
float dotp = 0.0;
for(i=0; i < 1024; i++)
{
   dotp += a[i]*b[i];
}
\end{verbatim}
To compile this code, we use the clang-2.8 \cite{clang} C compiler, 
which is used to emit LLVM \cite{llvm} byte-code.  The LLVM byte-code
is transformed through a series of steps by our compiler tools to
produce the final hardware, which is depicted in Figure \ref{fig:dotP}.
\begin{figure*}[ht]
  \centering
  \includegraphics[width=15cm]{dotP.eps}
  \caption{Control-data-storage virtual circuit model.}
  \label{fig:dotP}
\end{figure*}
The virtual circuit in Figure \ref{fig:dotP} has three components,
described below.

@@

Some data-path operations (such as the multiplexor
shown on the top and the decision operation shown at the bottom
left in Figure \ref{fig:dotP} follow a simpler protocol.  The multiplexor
has a pair of requests and a single acknowledge, with the condition
that at most one of the requests is received at any time instant.
The input corresponding to the request is then sampled and stored
in the output net of the multiplexor.
The decision operation has a single request and two acknowledes.  Upon
receipt of the request symbol, the decision operation checks its input net
and emits one of the two acknowledges depending on whether the input
is zero/non-zero.

@@

Note that the data-path only shows the operations and their interconnection.
When the data-path is implemented as hardware, multiple operations may
be mapped to a single operator depending on cost/performance tradeoffs.  When
this is done, multiplexing logic has to be introduced in the hardware.  These
decisions and manipulations are performed in the compiler stage which is
responsible for transforming the virtual circuit to VHDL.

@@

%% 
%% Petri-net, constructed according to
%% certain rules.  Req/Ack transitions
%%
The control-path in the virtual circuit encodes
all the sequencing that is necessary for correct
operation of the assembly.
The control-path (shown on the left
in Figure \ref{fig:dotP}) is modeled as a Petri-net with
a unique entry point and a unique exit point.  The Petri-net
is constructed using a set of production rules which guarantee
liveness and safeness \cite{ahir}.  Transitions in the Petri-net
are associated with output symbols to the data-path (these can
be described by the regular expressions *sr and *cr)
and input symbols from the data-path (these are of the form *sa and
*ca).  The *sr symbols instruct an element in the data-path to
sample its inputs and the *cr symbols instruct an element in the
data-path to update its outputs (all outputs of data-path elements
are registered).  The *sa and *ca symbols are acknowledgements
from the data-path which indicate that the corresponding requests
have been served.

@@

In our proposed, loop pipelining scheme, we use the
following ordering scheme.
If A is an operation in the loop body,
denote the
$k^{th}$ execution of A by $A_k$.  Since each operation
has  events sr, sa, cr, ca, we denote these by
$A_k.sr$, $A_k.sa$, $A_k.cr$, $A_k.ca$ respectively. We impose 
the following dependency rules on operations across
loop iterations.
\begin{itemize}
\item $A_k.sa \rightarrow A_{k+1}.sr$ for all operations A: that is,
the next execution of A cannot start until the current execution has
finished sampling the inputs.
\item $A_k.ca \rightarrow A_{k+1}.cr$ for all operations A: that is,
the completion of the next execution of A can be initiated only
after the current execution of A has completed.
\item If $A \rightarrow B$ is a RAW dependency, then 
$B_k.sa \rightarrow A_{k+1}.cr$.  That is, until B has
sampled the result of the current A, the next completion
of A cannot start.
\item If $A \rightarrow B$ is a WAR dependency, then
$B_k.ca \rightarrow A_{k+1}.sr$.  That is, the next
A cannot start until the current B has completed.
\item If P, Q are successive load/stores, with at least
one of them being a store, then 
$Q_k.sa \rightarrow P_{k+1}.sr$.  That is, the next P
cannot start until the current Q has acknowledged that it
has started.
\end{itemize}
The mechanism for incorporating RAW and WAR dependencies is
illustrated in Figure \ref{fig:pipelineMechanism}
for RAW and WAR dependencies within the loop body.
The reverse dotted arc is a marked arc (it initially
carries a single token).
\begin{figure}[ht]
  \centering
  \includegraphics[width=7cm]{pipelineMechanism.eps}
  \caption{Control-path mechanism for handling RAW, WAR dependencies for loop-pipelining.}
  \label{fig:pipelineMechanism}
\end{figure}

>>>>

\author{Author's name Suppressed for Blind Review}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

@@

In our work, we have developed a set of compiler tools which take a C-description of
an algorithm (with the only important restrictions being the requirement that
cycles are not present in the call graph of the source program, and the lack of 
support for {\bf function} pointers, normal pointers are supported), and produces
a VHDL description which is provably equivalent to the original program \cite{SameerPhD}.
Results on the application of this compiler to single-threaded programs show
that the circuits (when synthesized to VLSI layout) are substantially more
energy efficient than a processor, but that the performance of the processor
is usually better than that of the synthesized circuit \cite{ahirDsd2010}.
If multi-threaded source programs are used as a starting point, the circuits produced 
are quite competitive with hand-designed RTL \cite{ahirUsenix2012}.
Quantitatively, the results in \cite{ahirDsd2010} and \cite{ahirUsenix2012} indicate
that the gap between the synthesized hardware and modern
pipelined processors (for single-threaded
programs) and between the synthesized hardware and hand-crafted hardware (for
multi-threaded programs) is of the order of 2-3X.  That is, the performance
of the synthesized hardware needs to improve by 2-3X in order to close
the respective performance gaps.

@@

In the context of compilation to a pipelined processor, 
several loop optimizations
have been considered in literature, such as loop-unrolling, loop-peeling,
software-loop-pipelining etc. \cite{Wolfe}, \cite{Muchnick}, with the intent of
these optimizations being the extraction of as much
parallelism as possible from the single-threaded source program.

%% Algorithm-to-hardware:  can similar techniques work?
Similar loop-optimization techniques have been explored and
reported in the literature related to reconfigurable hardware (for
example \cite{WeinhardtLuk}, \cite{Cardoso}, \cite{KastnerPhd}).  For instance,
in the work reported by \cite{WeinhardtLuk},
loop optimizations are done in a manner analogous to the static
techniques used in software compilers, in which index 
expressions which depend on the induction variable
are analysed to identify dependencies and schedule operations
across iterations of the loop body.  An explicitly timed controller is synthesized
for the pipeline.  Another approach which works in a similar
manner is described in \cite{Cardoso}.  These approaches rely on a static analysis of
the loop, and the cases to which the approach can be applied
are restricted (but are still sufficiently general for most linear
algebra and digital signal processing kernels).

@@

We give a brief description of our compiler flow.  The
details are not relevant to this paper, and the interested
reader can find them in \cite{ahirDsd2010}.

@@

The compiler flow has been characterized over a wide variety
of applications \cite{ahirDsd2010}.   The XXXX\footnote{name suppressed for blind review.} tool-chain \cite{ahirUsenix2012} 
uses this compiler flow to map modular network appliances written in C++ to
an FPGA platform, and achieved about half the
performance possible using equivalent hand designed hardware.  
Thus,  the flow can be applied to ``real'' software
to produce competitive results.  The optimizations indicated in this
paper, should in principle, make the flow even more competitive.

@@

The virtual circuit generated by our compiler consists of three
cooperating components: the control-path, the data-path and
the storage system \cite{ahirDsd2010}.

To illustrate the model, we consider a simple example.
\begin{verbatim}
float a[1024], b[1024];
float dotp = 0.0;
for(i=0; i < 1024; i++)
{
   dotp += a[i]*b[i];
}
\end{verbatim}
To compile this code, we use the clang-2.8  C compiler, 
which is used to emit LLVM byte-code.  The LLVM byte-code
is transformed through a series of steps by our compiler tools to
produce a virtual circuit, which is depicted in Figure \ref{fig:dotP}.
\begin{figure*}[ht]
  \centering
  \includegraphics[width=15cm]{dotP.eps}
  \caption{Control-data-storage virtual circuit model.}
  \label{fig:dotP}
\end{figure*}
The virtual circuit in Figure \ref{fig:dotP} has three components,
described below.

@@

Some data-path operations (such as the multiplexor
shown on the top and the decision operation shown at the bottom
left in Figure \ref{fig:dotP}) follow a simpler protocol.  The multiplexor
has a pair of requests and a single acknowledge, with the condition
that at most one of the requests is received at any time instant.
The input corresponding to the request is then sampled and stored
in the output net of the multiplexor.
The decision operation has a single request and two acknowledes.  Upon
receipt of the request symbol, the decision operation checks its input net
and emits one of the two acknowledges depending on whether the input
is zero/non-zero.

@@

\noindent
{\bf Remark}
Note that the data-path only shows the operations and their interconnection.
When the data-path is implemented as hardware, multiple operations may
be mapped to a single operator depending on cost/performance tradeoffs.  When
this is done, multiplexing logic is introduced in the hardware.  These
decisions and manipulations are performed in the compiler stage which is
responsible for transforming the virtual circuit to VHDL.

@@

%% 
%% Petri-net, constructed according to
%% certain rules.  Req/Ack transitions
%%
The control-path in the virtual circuit encodes
all the sequencing that is necessary for correct
operation of the assembly.
The control-path (shown on the left
in Figure \ref{fig:dotP}) is modeled as a Petri-net with
a unique entry point and a unique exit point.  The Petri-net
is constructed using a set of production rules which guarantee
liveness and safeness \cite{ahirDsd2010}.  Transitions in the Petri-net
are associated with output symbols to the data-path (these can
be described by the regular expressions *sr and *cr)
and input symbols from the data-path (these are of the form *sa and
*ca).  The *sr symbols instruct an element in the data-path to
sample its inputs and the *cr symbols instruct an element in the
data-path to update its outputs (all outputs of data-path elements
are registered).  The *sa and *ca symbols are acknowledgements
from the data-path which indicate that the corresponding requests
have been served.

@@

In our dynamic loop pipelining scheme, we use the
following ordering scheme.
If A is an operation in the loop body,
denote the
$k^{th}$ execution of A by $A_k$.  Since each operation
has  events sr, sa, cr, ca, we denote these by
$A_k.sr$, $A_k.sa$, $A_k.cr$, $A_k.ca$ respectively. We impose 
the following dependency rules on operations across
loop iterations.
\begin{itemize}
\item $A_k.sa \rightarrow A_{k+1}.sr$ for all operations A: that is,
the next execution of A cannot start until the current execution has
finished sampling the inputs.
\item $A_k.ca \rightarrow A_{k+1}.cr$ for all operations A: that is,
the completion of the next execution of A can be initiated only
after the current execution of A has completed.
\item If $A \rightarrow B$ is a RAW dependency, then 
$B_k.sa \rightarrow A_{k+1}.cr$.  That is, until B has
sampled the result of the current A, the next completion
of A cannot start.
\item If $A \rightarrow B$ is a WAR dependency, then
$B_k.ca \rightarrow A_{k+1}.sr$.  That is, the next
A cannot start until the current B has completed.
\item If P, Q are successive load/stores, with at least
one of them being a store, then 
$Q_k.sa \rightarrow P_{k+1}.sr$.  That is, the next P
cannot start until the current Q has acknowledged that it
has started.
\end{itemize}
The mechanism for incorporating RAW and WAR dependencies is
illustrated in Figure \ref{fig:pipelineMechanism}
for RAW and WAR dependencies within the loop body.
The reverse dotted arc is a marked arc (it initially
carries a single token).
\begin{figure}[ht]
  \centering
  \includegraphics[width=7cm]{pipelineMechanism.eps}
  \caption{Control-path mechanism for handling RAW, WAR dependencies for loop-pipelining.}
  \label{fig:pipelineMechanism}
\end{figure}
